---
title: 'TDI Section 1: Project proposal'
author: "Maria Ayala"
date: "10/05/2021"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readxl)
library(tidyr)
library(tidyverse)
library(usmap)
library(dplyr)
```


For the TDI project, I'd like to propose an analysis of student educational outcomes. In recent sociological studies of economic stratification and intergenerational mobility, it has now been shown that the best predictor
of success in life is education. Looking very briefly at how states allocate and spend on education (https://www.urban.org/), we see a large disparity between key regions. And not only does this have immediate repercussions, but this affects key metrics that measure quality of life over time, such as upward mobility.

```{r spending, echo=FALSE}


spending <- read_excel("State-by-State Spending on Kids.xlsx", sheet = 1)
statepops <- read.csv("acs2017_census_tract_data.csv", header = TRUE)

statepops <- statepops %>%
  group_by(State) %>%
  summarise(popn = sum(TotalPop, na.rm = TRUE)) %>%
  filter(State != "Puerto Rico")

orange <- "#C9592E"

spending$fips <- fips(spending$state) # get geo IDs

spending <- left_join(spending,statepops, by = c("state" = "State"))
spending$spending2016_percapita <- spending$`2016`/spending$popn # normalize

plot_usmap(data = spending, values = "spending2016_percapita",  color = orange, labels=FALSE) + 
  scale_fill_continuous( low = "white", high = orange, 
                         name = "Spending", label = scales::comma
  ) + 
  theme(legend.position = "right") + 
  theme(panel.background = element_rect(colour = "black")) + 
  labs(title = "State spending on education per capita", caption = "Source: @ayalamars")
```
Given these disparities, it's vital that efforts are allocated appropriately. Developing a live and publicly accessible web-app that visualizes these key disparities will be immediately beneficial to efforts 
in deploying additional services (such as independent contractors that supplement education, some SaaS companies come to mind), as well as increase efforts in those affected regions. 

Using the publicly available Civil Rights Data Collection Datasets (https://www2.ed.gov/about/offices/list/ocr/data.html) we can for instance examine disparities in retention between demographic characteristics, such as race:
```{r retention, echo=FALSE}

retention <- read_excel("Retention-in-Grade-12.xlsx", sheet = 2)

ret_states <- retention[-1,] # remove totals for easy plotting

ret_states$fips <- fips(ret_states$state)# geo tags

ret_states <- left_join(ret_states,statepops, by = c("state" = "State"))
ret_states$HispRetpercapita <- as.numeric(ret_states$hispanic_num)/ret_states$popn # normalize

blue <- "#0000CD"

plot_usmap(data = ret_states, values = "hispanic_per",  color = blue, labels=FALSE) + 
  scale_fill_continuous( low = "white", high = blue, 
                         name = "Retention", label = scales::comma
  ) + 
  theme(legend.position = "right") + 
  theme(panel.background = element_rect(colour = "black")) + 
  labs(title = "Hispanic or Latino Student retention", caption = "Source: @ayalamars")

```
Or even target enrollment information for Calculus for underrepresented groups that can later be compared to the prevalence to the entire US population.
```{r calculus, echo=FALSE}
calc_states<- read_excel("Enrollment-in-Calculus.xlsx", sheet = 4)

calc_states$fips <- fips(calc_states$state)# geo tags

calc_states<- left_join(calc_states,statepops, by = c("state" = "State"))
calc_states$enrollpercap<- as.numeric(calc_states$black_num)/calc_states$popn # normalize

green <- "#228B22"

plot_usmap(data = calc_states, values = "enrollpercap",  color = green, labels=FALSE) + 
  scale_fill_continuous( low = "white", high = green, 
                         name = "Enrollment", label = scales::comma
  ) + 
  theme(legend.position = "right") + 
  theme(panel.background = element_rect(colour = "black")) + 
  labs(title = "Enrollment in Calculus for Black Females", caption = "Source: @ayalamars")

```

Together, these findings will help improve student outcomes by improving access to data so that the right services can be targeted to the right demographic, and ultimately improve graduation rates.

